Gradual Soundness: Lessons from Static Python

### 2021-12-16 pj-7.1/paper.tex
pj: type dynamic

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
Relative to these sources, the \SP{} takes some interesting turns
because of the engineering context at Instagram.
The following contextual points are important to bear in mind as we present
notable aspects of the type system:
\begin{itemize}
  \item
    Performance is the bottom line.
    Unless \SP{} helps production code run faster, the team will be disbanded
    and reallocated.
    %% 2021-12-14: too extreme? and what about soundness?
  \item
    Instagram is / had been using Cython for performance.
    Cython compiles Python source code to C extension modules for
    a faster Python [CITE].
    Instagram had converted several performance-critical modules to Cython prior
    to starting work on \SP{}.
    \SP{} replaced them.
    (How does this help understanding type system highlights?)
  \item
    The Instagram codebase is already typed.
    Developers use Pyre [CITE], an optional type checker for Python,
    to write type annotations and to drive IDE tools (e.g., autocompletion).
  \item
    Q. why is the python so nominal? style guide? java background?
\end{itemize}

\subsubsection{Type Dynamic}

%% 2021-12-14:
%% - anything to say about Any being a "meta" type in Python typing module?

\SP{} includes a dynamic type to accommodate untyped Python code.
Any un-annotated value or expression gets the dynamic type by default.
\SP{} does not make any predicitions / optimizations about dynamic code;
it assumes nothing and lets the Python runtime handle.

- lack of ann
- or, incorrect ann
- at toplevel, everything compatible
- diffirent rules inside data structures

The dynamic type is an important part of Static Python.
Dynamic avoids the need for whole-sale migration,
lets programmers work around dynamic features like splatting,
and is in some cases necessary for optimal performance (evidence?).

Can't really write the dynamic type.
But ANY unbound id. or non-type object gets interpreted as the dynamic type for now.

Cannot appear everywhere ... ChkDict can't hold a Dynamic (or rather, you can declare this type and cannot instantiate it),
and primitives are incompatible with Dynamic.

Dynamic behavior is always (?) the same as normal Python.

\subsubsection{Checked Types and Shallow Types}

@@

@@

  \item \textbf{First-order Typed Classes}.
    In order to benefit from types, class definitions must be defined at the top level of a module.
    Nested classes are supported by the language, but have type \code{Dynamic} according
    to the type checker.

  \item \textbf{Callable}.
    The type system cannot yet express first-class functions.
    The team plans to lift this restriction~(\cref{s:future}),
    but there is little internal pressure to do so.

@@

@@

\section{Design Principles}

@@

@@

\section{Conclusion}
\label{s:conclusion}

>>>>

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
%% - what are the idiomatic base SP types ... int or Int ?

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
Relative to these, the \SP{} takes some interesting turns
because of the engineering context at Instagram.
The following contextual points are important to bear in mind as we present
notable aspects of the type system:
\begin{itemize}
  \item
    Performance is the bottom line.
    Unless \SP{} helps production code run faster, the team will be disbanded
    and reallocated.
    %% 2021-12-14: too extreme? and what about soundness?
  \item
    %% Q. why is the python so nominal? style guide? java background?
    The codebase makes heavy use of first-order classes and objects.
    First-class classes, objects, and even functions are rare.
  \item
    Instagram is / had been using Cython to improve the performance of select modules.
    Cython compiles Python source code to C extension modules for
    a faster Python [CITE].
    Instagram had converted several performance-critical modules to Cython prior
    to starting work on \SP{}.
    \SP{} replaced them.
    (How does this help understanding type system highlights?)
  \item
    The Instagram codebase is already typed.
    Developers use Pyre [CITE], an optional type checker for Python,
    to write type annotations and to drive IDE tools (e.g., autocompletion).
\end{itemize}

\subsubsection{Type Dynamic}
\label{s:type-dynamic}

%% 2021-12-14:
%% - anything to say about Any being a "meta" type in Python typing module?
%% - 3 classes of code ... how does dynamic differ from untyped?

Like most gradual languages, \SP{} includes a dynamic type
that allows untyped expressions within a statically-typed context.
Whenever an expression or variable lacks a type annotation,
\SP{} uses the dynamic type as a default and skips most compile-time checks.
Every untyped Python program is thus a well-formed \SP{} program.
Furthermore, programmers can omit the annotations for specific
function arguments without losing the benefits of other type checks.
Freedom to omit annotations helps programmers convert modules to \SP{}
with little initial refactoring.

In addition to serving as a default, the dynamic type also replaces any undefined types.
For example, suppose that the type \code{X} is undefined in \SP{} and that a
program contains the annotation \code{ChkDict[int, X]}, which describes a
dictionary with integer keys and \code{X}-typed values.
\SP{} interprets this annotation as \code{ChkDict[int, dynamic]} and allows any sort
of values inside the dictionary at runtime.
To an outsider, this behavior might seem odd: why not reject undefined types with an error?
The reason is to reduce friction with the Pyre annotations that are already
widely-used at Instagram.
Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
efficiently, e.g., the type \code{Union[t1, ..., tn]} of arbitrary-width unions.
By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.

The dynamic type is not, however, the flexible dynamic type provided
by \emph{true} gradual languages~\cite{svcb-snapl-2015}.
Replacing part of a type with dynamic can lead to both compile-time errors
and run-time errors.
In other words, \SP{} satisfies neither the static nor the dynamic gradual guarantee.\footnote{We
assume a standard type precision relation. Of course, one could argue that these programs
do not break the gradual guarantees if type precision does not allow our uses of the dynamic type.}
\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
\begin{itemize}
  \item
    \cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
    The subclass definition raises a compile-time error because it attempts to override
    the typed \code{m1} method with another method that returns the dynamic type.
  \item
    \cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
    uses the type \code{ChkDict} for concrete-typed dictionaries (\cref{s:checked-type}).
    It sends a dictionary with integer values to a function that expects
    dictionaries with dynamic values.
    At runtime, the function rejects this argument because its type is not an
    exact match.
\end{itemize}
In sum, these behaviors show that performance is paramount in \SP{}.
Static errors in dynamic-typed code are designed to protect the integrity of
type-based optimizations.
%% other examples?
The dynamic errors are a consequence of strict run-time constraints, which provide stronger guarantees
and are simpler to validate than lax constraints.
%% hmph, unclear. Spell it out!

\begin{figure}
  \begin{subfigure}[t]{0.5\columnwidth}
  \begin{verbatim}
    class A:
      def m1(self)->int:
        return 0

    class B(A):
      def m1(self):
        return 0
    # Error: type dynamic does not match int
  \end{verbatim}
    \caption{Removing a type in \code{B} breaks the static gradual guarantee}
    \label{f:gg-failure-stat}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.5\columnwidth}
  \begin{verbatim}
    from __static__ import ChkDict

    def f(x: ChkDict[str, dyn]):
      return x["A"]

    d = ChkDict[str, int]({"A": 1})
    f(d)
    # Error: f expected ChkDict[str, dyn]
  \end{verbatim}
    \caption{Removing part of the type for \code{x} breaks the dynamic gradual guarantee}
    \label{f:gg-failure-dyn}
  \end{subfigure}
  \caption{\SP{} values performance over the gradual guarantees.}
  \label{fig:gg-failure}
\end{figure}

\subsubsection{Concrete Types and Shallow Types}
\label{s:checked-type}

@@

%% TODO very important, one of the main things IG does use!!!

@@

%  \item \textbf{First-order Typed Classes}.
%    In order to benefit from types, class definitions must be defined at the top level of a module.
%    Nested classes are supported by the language, but have type \code{Dynamic} according
%    to the type checker.
%
%  \item \textbf{Callable}.
%    The type system cannot yet express first-class functions.
%    The team plans to lift this restriction~(\cref{s:future}),
%    but there is little internal pressure to do so.

@@

%% TODO built-in vs. user defined ... 
%% Carl: I don't think this is quite true the way it's phrased here? in general
%%   for SP classes we wrap non-static subclass override methods at runtime (in our
%%   vtable implementation) to enforce correct return type, so that we don't need
%%   exact types, we can trust that Liskov is not broken.
%%     The cases where we tend to be limited to exact types is in optimizing
%%   operations on builtin types, where they aren't really static types and
%%   methods called on them aren't going through SP vtables, so we don't have
%%   the vtable wrappers to enforce LSP, we just know the behavior of the exact
%%   builtin type

%% TODO wherever we discuss optimizations, also talk about de-slowdowns
%% Carl: not sure where this belongs exactly, but we go to a lot of work in the
%%   compiler to intelligently minimize runtime checks by only inserting them when
%%   the source value is of dynamic type (this includes also skipping CHECK_ARGS if
%%   the function was statically invoked by another static function and thus we
%%   know the compiler checked argument types)

@@

%% 2021-12-15: C&D may have an SP off/on switch soon
%% - accumulated CPU savings since July 1 = ~2% total IG production CPU
%%   ... how does that number work?
%% - in the beginning (?) H1 got 1% from first conversions + 0.7% from all Cython (higher bar)
%%   - no Cython regressions, SP + JIT is good enough
%%   - some Cython still exists, imported as library code from another codebase
%% - rough measure: 0 to 25% per module
%%   ... but depends on nature of module
%%   ... doesn't measure the cooperative effect of static-to-static cross-module calls

%% - 530 SP modules
%%   - 400 from a codegen tool
%%   - 130 by hand
%%   - currently +~3/4 per week by hand

%% - pushback one time about removing MI
%% - otherwise, people happy with low-cost improvements

%% - review of 30 winning diffs "provided the bulk of our wins this half"
%%   most common changes, not in careful order, are:
%%   1) adjustments to tests to pass correct types instead of mock objects, as as
%%      not to fail type checks in SP code. This is an inherent and desired
%%      incompatibility; next half we plan to work on improving the Python mock
%%      framework to automatically create mocks of the correct types.
%%      - bg: lookup the dino email about mocks
%%   2) adjustments to mock assertions in tests about calls to adjust for the
%%      fact that SP currently converts all call arguments to be positional (we plan
%%      to fix this incompatibility by falling back to the original call arguments
%%      when we detect the target function has been patched.)
%%      - bg: ??? bit fuzzy about positional and others
%%   3) adjustments around class attributes vs instance attributes (Python allows
%%      this to be a bit fuzzy, where a class attribute can be a fallback default
%%      for an instance attribute if not set, but SP currently requires an attribute
%%      to be clearly one or the other, and we have more efficient access for
%%      instance attributes via LOAD_FIELD that we don’t have for class attributes.)
%%      - related to github #37 ?
%%   4) convert @classmethod that don’t actually need the class to @staticmethod,
%%      allowing for emitting direct INVOKE_FUNCTION instead of INVOKE_METHOD
%%   5) extract special types that our compiler doesn’t yet support/understand
%%      (classes decorated with @dataclass or inheriting from `enum.Enum`) out of
%%      the module we are converting to static. We have work in progress on
%%      supporting some Enum types, and plan to add a @dataclass intrinsic soon.
%%   6) Removing the use of super-dynamic Python features like `__setattr__` that
%%      we don’t support. So far in the cases where we’ve done this it hasn’t
%%      required a major rewrite.
%%      - bg: more details pls ,,, sketch the rewrite
%%   7) Removing the use of keyword-only arguments (bare `*`) in function
%%      signatures, since we don’t yet support calling such functions. We plan to
%%      support this next half.
%%   8) Convert Python integers and bools used in very hot code paths into SP
%%      machine primitives.
%%      - bg: focus on block, convert, no problem?
%%   9) Occasionally converting a regular Python list or dict to a CheckedList or
%%      CheckedDict.
%%      - bg: how occasional? worried about power of the "gradual soundness" claim
%%
%% common theme: gotta change more tests than code
%%
%% significant change: replaced `contextlib.ContextDecorator` with `__static__.ContextDecorator`
%%  removes an extra call layer
%%  helps with (timing [ context managers / decorators ])
%%
%% all of this is the low-hanging fruit, haven't tackled uses of metaclasses even though looks promising
%%  because the pay / payoff isn't yet attractive

@@

%% Immediate future: fix adoption friction points
%% 2021-12-15: WHAT ARE THESE?

%% engineering: JIT profiling, instead of hand-requested JIT list
%% model: metaclasses

@@

\section{Conclusion / Lessons / Design Principles}
\label{s:conclusion}
